---
title: "Getting started with normScore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with normScore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5
)
```


## Overview

`normScore` provides tools to evaluate and rank normalization methods for
omics data using a composite multi-metric scoring framework.

The package is designed to help compare different normalization strategies
based on several complementary criteria, including variability, correlation
structure, MA-trend behavior, relative log expression consistency, and
sample-level intensity consistency.

In this vignette, we show a minimal workflow using simulated data.

## Simulating example data

To keep the package lightweight and reproducible, `normScore` includes a
data simulation function that generates proteomics-like two-group datasets.

```{r simulateData}
library(normScore)

simData <- simulateData(
    nProteins = 500,
    nPerGroup = 4,
    sampleShiftSd = 0.12,
    sampleShiftCap = 1,
    sampleSdStrength = 1.2,
    sampleSdRho = 0.3,
    rhoWithin = 0.6,
    rhoBetween = 0.3,
    loadingSd = 0.45,
    sigmaHi = 0.03,
    sigmaLo = 0.4,
    gammaSigma = 3.8,
    kMnar = 1.6,
    addMissing = FALSE,
    seed = 123
)

```

The simulated object contains three main elements:

```{r justSimDataStr}
names(simData)
```

* `logData`: simulated data on the log2 scale
* `rawData`: simulated data on the raw scale
* `metadata`: sample annotation with group membership

We can inspect the sample metadata:

```{r simMetadata}
head(simData$metadata)
```

## Building a list of normalized datasets

The main function of the package, `normScore()`, expects a named list of
normalized datasets. Each element of the list should be a matrix or data
frame with the same rows (features) and columns (samples).

For illustration purposes, we create a few simple variants of the simulated
log-scale data. In a real analysis, these would correspond to different
normalization methods.

```{r normalization}
# Median normalization
MedianMat <- NormalyzerDE::medianNormalization(simData$rawData)
colnames(MedianMat) <- colnames(simData$rawData)
rownames(MedianMat) <- rownames(simData$rawData)

# Quantile normalization
QuantileMat <- limma::normalizeQuantiles(
    simData$logData
)
colnames(QuantileMat) <- colnames(simData$logData)
rownames(QuantileMat) <- rownames(simData$logData)


# CyclicLoess normalization
CyclicNorm <- limma::normalizeCyclicLoess(
    simData$logData, method="fast",  adaptive.span=FALSE
)


# All together
normalizedDataList <- list(
    Log = simData$logData,
    Median = MedianMat,
    Quantile = QuantileMat, 
    CyclicLoess = CyclicNorm
)

names(normalizedDataList)
```

## Running `normScore()`

We now compute the normalization ranking.

```{r normScoreUsage}
result <- normScore(
    normalizedDataList = normalizedDataList,
    groupData = simData$metadata,
    rawData = simData$rawData,
    returnDetails = TRUE,
    nBoot = 150
)
```

## Final ranking

The `finalRanking` element contains the final composite score for each
normalization method.

```{r finalNormScoreRank}
result$finalRanking
```

Lower values indicate better normalization performance according to the
scoring framework implemented in the package.

## Detailed scores

The `detailRanking` element contains the scaled item-wise scores used to
build the final ranking.

```{r detailedRanking}
result$detailRanking
```

The six item scores correspond to:

1. pooled coefficient of variation
2. within-group sample correlation
3. MA-plot trend deviation
4. mean-SD trend deviation
5. RLE consistency
6. total intensity consistency

These are scaled and combined into a final score.

## Bootstrap summary

If `returnDetails = TRUE`, the function also computes bootstrap-based
summary scores and confidence intervals.

```{r strSimulatedData}
result$bootstrapScore
```

This output contains:

* the normalization method name
* the mean bootstrap normScore
* the lower 95% confidence bound
* the upper 95% confidence bound

## Bootstrap plot

The bootstrap summary can also be visualized directly:

```{r plotBootstrapInterval}
plotBootstrapNormScore(result)
```

## Diagnostic plots {.tabset}

Individual diagnostic plots from state-of-art normalization criteria can be
visualed with the following function:

```{r getDiagnosticPlots}
allPlots <- plotNormScoreDiagnostics(
    normalizedDataList = normalizedDataList,
    groupData = simData$metadata,
    rawData = simData$rawData
)

names(allPlots)

# Plotting each in a different tab
```


```{r diagnPlots, echo = FALSE, results = "asis", fig.height=10, fig.width=8}
for (individualPlot in names(allPlots)) {
    cat("\n### ", individualPlot, " {-}\n\n")
    print(allPlots[[individualPlot]])
    cat("\n\n")
}
```

## Notes on the scoring framework

The `normScore()` function combines multiple complementary criteria into a
single ranking. This is useful because no single metric fully captures
normalization quality.

Some items emphasize within-group consistency, whereas others evaluate
distributional behavior or trend distortions introduced by normalization.

The resulting score should therefore be interpreted as a comparative measure
across candidate normalization methods rather than as an absolute quality
index.

## Important considerations

Before running `normScore`, verify that:

- all assays contain the same rows and columns;
- feature and sample names are unique;
- the order of `groupData$Samples` matches the assay columns;
- missing values are represented consistently across assays;
- every sample has a valid group assignment.

## Session information

```{r versions}
sessionInfo()
```
