---
title: "Getting started with normScore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{normScore-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```


## Overview

`normScore` provides tools to evaluate and rank normalization methods for
omics data using a composite multi-metric scoring framework.

The package is designed to help compare different normalization strategies
based on several complementary criteria, including variability, correlation
structure, MA-trend behavior, relative log expression consistency, and
sample-level intensity consistency.

In this vignette, we show a minimal workflow using simulated data.

## Simulating example data

To keep the package lightweight and reproducible, `normScore` includes a
data simulation function that generates proteomics-like two-group datasets.

```{r}
library(normScore)

simData <- simulateData(
  nProteins = 500,
  nPerGroup = 4,
  addMissing = FALSE,
  seed = 123
)
```

The simulated object contains three main elements:

```{r}
names(simData)
```

* `logData`: simulated data on the log2 scale
* `rawData`: simulated data on the raw scale
* `metadata`: sample annotation with group membership

We can inspect the sample metadata:

```{r}
head(simData$metadata)
```

## Building a list of normalized datasets

The main function of the package, `normScore()`, expects a named list of
normalized datasets. Each element of the list should be a matrix or data
frame with the same rows (features) and columns (samples).

For illustration purposes, we create a few simple variants of the simulated
log-scale data. In a real analysis, these would correspond to different
normalization methods.

```{r}
normalizedDataList <- list(
  Log = simData$logData,
  Shifted = simData$logData + 0.1,
  Scaled = simData$logData * 1.01
)

names(normalizedDataList)
```

## Running `normScore()`

We now compute the normalization ranking.

```{r}
result <- normScore(
  normalizedDataList = normalizedDataList,
  groupData = simData$metadata,
  rawData = simData$rawData,
  returnDetails = FALSE,
  nBoot = 50
)
```

## Final ranking

The `finalRanking` element contains the final composite score for each
normalization method.

```{r}
result$finalRanking
```

Lower values indicate better normalization performance according to the
scoring framework implemented in the package.

## Detailed scores

The `detailRanking` element contains the scaled item-wise scores used to
build the final ranking.

```{r}
result$detailRanking
```

The six item scores correspond to:

1. pooled coefficient of variation
2. within-group sample correlation
3. MA-plot trend deviation
4. mean-SD trend deviation
5. RLE consistency
6. total intensity consistency

These are scaled and combined into a final score.

## Bootstrap summary

If `onlyFinalRanking = FALSE`, the function also computes bootstrap-based
summary scores and confidence intervals.

```{r}
result$bootstrapScore
```

This output contains:

* the normalization method name
* the mean bootstrap normScore
* the lower 95% confidence bound
* the upper 95% confidence bound

## Bootstrap plot

The bootstrap summary can also be visualized directly:

```{r}
plotBootstrapNormScore(result)
```

## Notes on the scoring framework

The `normScore()` function combines multiple complementary criteria into a
single ranking. This is useful because no single metric fully captures
normalization quality.

Some items emphasize within-group consistency, whereas others evaluate
distributional behavior or trend distortions introduced by normalization.

The resulting score should therefore be interpreted as a comparative measure
across candidate normalization methods rather than as an absolute quality
index.

## Session information

```{r}
sessionInfo()
```
