---
title: "NormScore with SummarizedExperiment and NormalyzedDE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NormScore with SummarizedExperiment and NormalyzedDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = "center",
    warning = FALSE,
    message = FALSE
)
```

## Introduction

`SummarizedExperiment` is a standard Bioconductor container for storing
high-dimensional molecular data together with feature and sample metadata.
A single object can contain several assays, making it convenient for keeping
the original data and multiple normalized versions aligned.

This vignette illustrates how to:

1. create a `SummarizedExperiment` object;
2. store several normalized matrices as assays;
3. extract the matrices and sample groups required by `normScore`;
4. evaluate the normalization methods;
5. add the resulting scores back to the object metadata.

## Required packages

```{r packages}
library(normScore)
library(SummarizedExperiment)
```

The packages used only in this vignette should be included in the
`Suggests` field of the package `DESCRIPTION` file:

```text
Suggests:
    BiocStyle,
    knitr,
    rmarkdown,
    SummarizedExperiment
VignetteBuilder: knitr
```

## Example data

For illustration, we generate a small log-transformed abundance matrix using 
`simulateData` function from `normScore` data package.

Rows represent proteins and columns represent samples in logData and rawData 
items from the output `simData` object.

```{r example-data}
simData <- simulateData(
    nProteins = 500,
    nPerGroup = 4,
    sampleShiftSd = 0.12,
    sampleShiftCap = 1,
    sampleSdStrength = 1.2,
    sampleSdRho = 0.3,
    rhoWithin = 0.6,
    rhoBetween = 0.3,
    loadingSd = 0.45,
    sigmaHi = 0.03,
    sigmaLo = 0.4,
    gammaSigma = 3.8,
    kMnar = 1.6,
    addMissing = FALSE,
    seed = 456
)

rawData <- simData$rawData
```


## Build a SummarizedExperiment object

The original and normalized matrices are stored as separate assays.
Sample-level information is stored in `colData`, whereas optional
feature-level annotations can be stored in `rowData`.

```{r se-object}
sampleData <- S4Vectors::DataFrame(
    sample = simData$metadata$Samples,
    group = simData$metadata$Groups
)

se <- SummarizedExperiment::SummarizedExperiment(
    assays = list(rawData = rawData),
    colData = sampleData,
    metadata = list(
        sample = "sample",
        group = "group")
)
```

## Normalization using NormalyzerDE

Using normalyzerDE different normalization methods are simultaneously applied 
to the input data.

```{r normalize-data}
normalyzerObject <- NormalyzerDE::getVerifiedNormalyzerObject(
    jobName = "normScore_example",
    summarizedExp = se,
    noLogTransform = FALSE
)

normalyzerResults <- NormalyzerDE::normMethods(
    normalyzerObject,
    normalizeRetentionTime = FALSE
)
```


## Prepare normScore inputs

`normScore` evaluates a named list of matrices. We therefore extract the
normalized assays from the `SummarizedExperiment` object.

```{r prepare-normalized-list}
normalizedDataList <- methods::slot(
    normalyzerResults,
    "normalizations"
)
```

Setting rownames as some normalized matrices from normalyzerDE 
lost protein rownames:

```{r adjust-rownames}
normalizedDataList <- lapply(
    normalizedDataList,
    function(x) {
        rownames(x) <- rownames(rawData)
        x
    }
)
```

The group information is constructed from `colData`. The sample order must
match the matrix columns.

```{r prepare-group-data}
groupData <- data.frame(
    Samples = sampleData$sample,
    Groups = sampleData$group
)

stopifnot(
    identical(groupData$Samples, colnames(rawData)),
    all(vapply(
        normalizedDataList,
        function(x) identical(colnames(x), groupData$Samples),
        logical(1)
    ))
)

groupData
```

## Run normScore  {.tabset}

The extracted objects can now be supplied to `normScore`.

```{r run-normscore}
normScoreResults <- normScore(
    normalizedDataList = normalizedDataList,
    groupData = groupData,
    rawData = rawData
)
```

Depending on the final argument names of the exported `normScore()`
function, this call may need to be adapted to the current package API.

The returned object can be inspected directly:

```{r inspect-results}
names(normScoreResults)

normScoreResults$finalRanking
```

For example, the score or ranking table can be displayed after selecting
the corresponding component returned by the package:

```{r score-table}
normScoreResults$detailRanking
```

Similarly, diagnostic plots can be accessed from the plotting component:

```{r diagnostic-plots}
allPlots <- plotNormScoreDiagnostics(
    normalizedDataList = normalizedDataList, 
    groupData = groupData, 
    rawData = rawData
)
```


```{r plotDiagn, echo = FALSE, results = "asis", fig.height=10, fig.width=8}
for (individualPlot in names(allPlots)) {
    cat("\n### ", individualPlot, " {-}\n\n")
    print(allPlots[[individualPlot]])
    cat("\n\n")
}
```

## Store results in the SummarizedExperiment metadata

Analysis-level information that is not naturally associated with individual
features or samples can be stored in `metadata(se)`.

```{r store-results}
metadata(se)$normScore <- normScoreResults
```

The results can then be recovered without separating them from the data:

```{r recover-results}
storedResults <- metadata(se)$normScore
names(storedResults)
```

A compact summary table may also be stored separately:

```{r store-summary}
metadata(se)$normalizationRanking <- normScoreResults$score
```

## Add a selected normalized assay

Once a normalization method has been selected, its matrix can remain as an
assay of the same object. For example, if median normalization is selected:

```{r selected-assay}
assay(se, "Selected") <- normalizedDataList$Quantile

assayNames(se)
```

The selected assay can then be used in downstream Bioconductor workflows:

```{r downstream-example}
selectedData <- assay(se, "Selected")

dim(selectedData)
```

## Working with an existing SummarizedExperiment

For an existing object, the minimum workflow is:

```{r compact-workflow, eval=FALSE}
assayNames(se)
colnames(se)
colData(se)

rawData <- assay(se, "Unnormalized")

normalizationMethods <- c("Mean", "Median", "Quantile")

normalizedDataList <- lapply(
    normalizationMethods,
    function(method) assay(se, method)
)
names(normalizedDataList) <- normalizationMethods

groupData <- data.frame(
    Samples = colnames(se),
    Groups = as.character(colData(se)$group)
)

normScoreResults <- normScore(
    normalizedDataList = normalizedDataList,
    groupData = groupData,
    rawData = rawData
)

metadata(se)$normScore <- normScoreResults
```

## Session information

```{r session-info}
sessionInfo()
```
